Types from data: \textnormal{Making structured data first-class citizens in F\#}

### 2016-02-04 pldi-2016/paper.tex
Final edits (section 1 and 2)

\documentclass[10pt,preprint,clearpagebib]{sigplanconf}

@@

% TODO: Discuss 'stability' of the inference (minor change in schema does not cause dramatic cahnges)
% TODO: Say that variant is a last-resort thing
% TODO: Practical evaluation (lots of downloads, lots of Github - how often people customize samples?)
% TODO: Explain why we need option type (because T + null is a variant)
% TODO: In the PLDI version, we delete all the unnecessary runtime conversions
% but we should say that those happen for practical reasons (e.g. float -> int, or anyhing to string)

@@

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}
\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{20yy} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
%\doi{nnnnnnn.nnnnnnn}

@@

In this paper, we integrate external structured data into F\# programming. As most real-world data
does not come with an explicit schema, we develop a shape inference algorithm that infers a shape from 
representative sample documents and integrates it into the F\# type system using type providers.
We present a formalization and a relative type soundness theorem for a simplified version of this process.

@@

\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC"}) \\
 \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
 |~\ident{Record}(\ident{root})\rightarrow \\
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\
 \quad |~\ident{Record}(\ident{main})\rightarrow \\
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f!}~\ident{num} \\
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format} 
\end{array}
\end{equation*}
%
The code assumes that the response has a particular shape described in the documentation. The
root node must be a record with a \strf{main} field, which has to be another record containing
a numerical \strf{temp} field. When the shape is different, the code simply fails with an exception. 
While not immediately unsound, the code is manifestly prone to errors if strings are misspelled 
or incorrect shape assumed.

@@

\vspace{-0.1em}
\noindent
The supplementary screencast illustrates the practical developer experience using 
  F\# Data with JSON, XML and CSV.

@@

\begin{itemize}
\item The mechanism is predictable. The user directly works with the provided types and should 
  understand why a specific type was produced from a given sample.\footnote{In particular, we do 
  not use probabilistic methods where adding an additional sample could completely change the 
  shape of the type.}

@@

@@

The JSON format is a popular data exchange format based on 
JavaScript data structures. The following is the definition of \ident{JsonValue} 
used earlier (\S\ref{sec:introduction}) to represent JSON data:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{JsonValue} = \\[0.1em]
 \quad|~ \ident{Number}~\kvd{of}~\ident{float} ~~~~|~ \ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
 \quad|~ \ident{String}~\kvd{of}~\ident{string} ~~~~~\, |~ \ident{Null} \\[0.1em]
 \quad|~ \ident{Record}~\kvd{of}~\ident{Map}\langl\ident{string}, \ident{JsonValue}\rangl \\[0.1em]
 \quad|~ \ident{Array}~\kvd{of}~\ident{JsonValue}[] \\[0.1em]
\end{array}
\end{equation*}
%
The earlier example used only a nested record containing a number. To demonstrate other 
aspects of the JSON type provider, we look at an example that also involves an array:
%
{\small{
\begin{verbatim}
  [ { "name":"Jan", "age":25 }, { "name":"Tomas" },
    { "name":"Alexander", "age":3.5 } ]
\end{verbatim}
}}
%
\noindent
The standard way to print the names and ages would be to pattern match on the parsed 
\ident{JsonValue}, check that the top-level node is a \ident{Array} and iterate over the elements 
checking that each element is a \ident{Record} with certain properties. We would throw an exception 
for values of an incorrect shape. As before, the code would specify field names as strings, which 
is error prone and can not be statically checked.

@@

The \ident{JsonProvider} uses a shape inference algorithm and provides
the following F\# types for the sample:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Entity}~=  \\
 \quad \kvd{member}~\ident{Name}~:~\ident{string} \\
 \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{float}\rangl \\[0.5em]
 \kvd{type}~\ident{People}~=  \\
 \quad \kvd{member}~\ident{GetSample}~:~\ident{unit}~\rightarrow~\ident{Entity}[] \\
 \quad \kvd{member}~\ident{Parse}~:~\ident{string}~\rightarrow~\ident{Entity}[] \\
\end{array}
\end{equation*}
%
The type \ident{Entity} represents the person. The field \ident{Name} is available for all
sample values and is inferred as \ident{string}. The field \ident{Age} is marked as optional,
because the value is missing in one sample. The two age values are an integer $25$ and a 
float $3.5$ and so the common inferred type is \ident{float}.

@@

\paragraph{Error handling.}
In addition to the structure of the types, the type provider also specifies what code should be 
executed at run-time in place of \ident{item.Name} and other operations. The runtime behaviour is 
the same as in the earlier hand-written sample (\S\ref{sec:introduction}) -- a member access 
throws an exception if data does not have the expected shape.

@@

For this reason, we do not infer a closed choice between heading, paragraph and image. In the 
subsequent formalization, we introduce a \emph{top shape} (\S\ref{sec:inference-types}) and extend 
it with labels capturing the statically known possibilities (\S\ref{sec:inference-vars}). The 
\emph{labelled top shape} is mapped to the following type:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Element}~=  \\[0.1em]
 \quad \kvd{member}~\ident{Heading}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Paragraph}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Image}~:~\ident{option}\langl \ident{Image} \rangl\\
\end{array}
\end{equation*}
%
This provides access to the elements known statically from the sample. However the type is 
deliberately `weak', because the user needs to explicitly handle the case when a value is not 
a statically known element. The above code uses \ident{Option.iter} to skip all unknown elements.

@@

\subsection{Summary}
\label{sec:providers-sum}
Throughout the introduction, we used data sets that demonstrate the typical problems that are frequent
in the real-world (missing data, inconsistent encoding of primitive values and heterogeneous shapes). 
The following JSON response with government debt information returned by the World Bank\footnote{Available at 
\url{http://data.worldbank.org}} demonstrates all three problems:
%
{\small{
\begin{verbatim}
  [ { "page": 1, "pages": 5 },
    [ { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2012", "value": null },
      { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2010", "value": "35.14229" } ] ]
\end{verbatim}
}}
%
\noindent
First of all, the top-level element is a collection containing two values of different shape.
The first is a record with meta-data about the current page and the second is an array with data. 
The actual F\# Data implementation supports a concept of heterogeneous collections (briefly outlined
in \S\ref{sec:impl-hetero}) and provides a type with properties \ident{Record} for the former and 
\ident{Array} for the latter. Second, the field \ident{value} is \kvd{null} for some records. Third, 
numbers can be represented in JSON as numeric literals (without quotes), but here, they are 
returned as string literals instead.\footnote{This is often used to avoid non-standard numerical 
types of JavaScript.}

@@

\section{Formalising type providers}
\label{sec:formal}

@@

>>>>

\documentclass[10pt,clearpagebib]{sigplanconf}

@@

@@

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}
\conferenceinfo{PLDI '16}{June 13--17, 2016, Santa Barbara, CA, United States} 
\copyrightyear{2016} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
%\doi{nnnnnnn.nnnnnnn}

@@

We present F\# Data, a library that integrates external structured data into F\#. As most real-world 
data does not come with an explicit schema, we develop a shape inference algorithm that infers a 
shape from representative sample documents. We then integrate the inferred shape into the F\# type 
system using type providers. We formalize the process and prove a relative type soundness theorem.

@@

\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC"}) \\
 \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
 |~\ident{Record}(\ident{root})\rightarrow \\
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\
 \quad |~\ident{Record}(\ident{main})\rightarrow \\
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f!}~\ident{num} \\
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format} 
\end{array}
\end{equation*}
%
The code assumes that the response has a particular shape described in the documentation. The
root node must be a record with a \strf{main} field, which has to be another record containing
a numerical \strf{temp} field representing the current temperature. When the shape is different, 
the code fails. While not immediately unsound, the code is prone to errors if strings are 
misspelled or incorrect shape assumed.

@@

@@

\begin{itemize}
\item The mechanism is predictable (\S\ref{sec:impl-stable}). The user directly works with the 
  provided types and should understand why a specific type was produced from a given sample.

@@

\noindent
The supplementary screencast provides further illustration of the practical developer
experience using F\# Data.

@@

The JSON format is a popular data exchange format based on 
JavaScript data structures. The following is the definition of \ident{JsonValue} 
used earlier (\S\ref{sec:introduction}) to represent JSON data:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{JsonValue} = \\[0.1em]
 \quad|~ \ident{Number}~\kvd{of}~\ident{float} \\[0.1em]
 \quad|~ \ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
 \quad|~ \ident{String}~\kvd{of}~\ident{string}\\[0.1em]
 \quad|~ \ident{Record}~\kvd{of}~\ident{Map}\langl\ident{string}, \ident{JsonValue}\rangl \\[0.1em]
 \quad|~ \ident{Array}~\kvd{of}~\ident{JsonValue}[] \\[0.1em]
 \quad|~ \ident{Null} \\[0.1em]
\end{array}
\end{equation*}
%
The earlier example used only a nested record containing a number. To demonstrate other 
aspects of the JSON type provider, we look at an example that also involves an array:
%
{\small{
\begin{verbatim}
  [ { "name":"Jan", "age":25 }, 
    { "name":"Tomas" },
    { "name":"Alexander", "age":3.5 } ]
\end{verbatim}
}}
%
\noindent
The standard way to print the names and ages would be to pattern match on the parsed 
\ident{JsonValue}, check that the top-level node is a \ident{Array} and iterate over the elements 
checking that each element is a \ident{Record} with certain properties. We would throw an exception 
for values of an incorrect shape. As before, the code would specify field names as strings, which 
is error prone and can not be statically checked.

@@

The \ident{JsonProvider} uses a shape inference algorithm and provides
the following F\# types for the sample:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Entity}~=  \\
 \quad \kvd{member}~\ident{Name}~:~\ident{string} \\
 \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{float}\rangl \\[0.5em]
 \kvd{type}~\ident{People}~=  \\
 \quad \kvd{member}~\ident{GetSample}~:~\ident{unit}~\rightarrow~\ident{Entity}[] \\
 \quad \kvd{member}~\ident{Parse}~:~\ident{string}~\rightarrow~\ident{Entity}[] \\
\end{array}
\end{equation*}
%
The type \ident{Entity} represents the person. The field \ident{Name} is available for all
sample values and is inferred as \ident{string}. The field \ident{Age} is marked as optional,
because the value is missing in one sample. In F\#, we can then use \ident{Option.iter} to call
the specified function (printing) only when an optional value is available. The two age values 
are an integer $25$ and a float $3.5$ and so the common inferred type is \ident{float}.
The names of the properties are normalized to follow standard F\# naming conventions
as discussed later (\S\ref{sec:impl-naming}).

@@

\paragraph{Error handling.}
In addition to the structure of the types, the type provider also specifies the code of 
operations such as \ident{item.Name}. The runtime behaviour is 
the same as in the earlier hand-written sample (\S\ref{sec:introduction}) -- a member access 
throws an exception if data does not have the expected shape.

@@

For this reason, we do not infer a closed choice between heading, paragraph and image. In the 
subsequent formalization, we introduce a \emph{top shape} (\S\ref{sec:inference-types}) and extend 
it with labels capturing the statically known possibilities (\S\ref{sec:inference-vars}). The 
\emph{labelled top shape} is mapped to the following type:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Element}~=  \\[0.1em]
 \quad \kvd{member}~\ident{Heading}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Paragraph}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Image}~:~\ident{option}\langl \ident{Image} \rangl\\
\end{array}
\end{equation*}
%
\ident{Element} is an abstract type with properties. It can represent the statically known
elements, but it is not limited to them. For a table element, all three properties would return 
\ident{None}.

Using a type with optional properties provides access to the elements known statically from the 
sample. However the user needs to explicitly handle the case when a value is not a statically 
known element. In object-oriented languages, the same could be done by providing a class hierarchy, 
but this loses the easy discoverability when ``.'' is typed.

@@

\subsection{Real-world JSON services}
\label{sec:providers-sum}

Throughout the introduction, we used data sets that demonstrate the typical problems frequent
in the real-world (missing data, inconsistent encoding of primitive values and heterogeneous shapes). 
The government debt information returned by the World Bank\footnote{Available at 
\url{http://data.worldbank.org}} demonstrates all three:
%
{\small{
\begin{verbatim}
  [ { "pages": 5 },
    [ { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2012", "value": null },
      { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2010", "value": "35.14229" } ] ]
\end{verbatim}
}}
%
\noindent
First, the field \ident{value} is \kvd{null} for some records. Second, numbers in JSON can be 
represented as numeric literals (without quotes), but here, they are returned as string literals 
instead.\footnote{This is often used to avoid non-standard numerical types of JavaScript.}
Finally, the top-level element is a collection containing two values of different shape.
The record contains meta-data with the total number of pages and the array contains the data.
F\# Data supports a concept of heterogeneous collection (outlined in in \S\ref{sec:impl-hetero})
and provides the following type:
%
\begin{equation*}
\begin{array}{l}
\kvd{type}~\ident{Record}~=  \\
 \quad \kvd{member}~\ident{Pages}~:~\ident{int} \\[0.5em]
\kvd{type}~\ident{Item}~=  \\
 \quad \kvd{member}~\ident{Date}~:~\ident{int} \\
 \quad \kvd{member}~\ident{Indicator}~:~\ident{string} \\
 \quad \kvd{member}~\ident{Value}~:~\ident{option}\langl \ident{float}\rangl\\[0.5em]
\kvd{type}~\ident{WorldBank}~=  \\
 \quad \kvd{member}~\ident{Record}~:~\ident{Record}\\
 \quad \kvd{member}~\ident{Array}~:~\ident{Item}[]\\[0.5em]
\end{array}
\end{equation*}
%
The inference for heterogeneous collections infers the multiplicities and shapes of nested 
elements. As there is exactly one record and one array, the provided type \ident{WorldBank} exposes 
them as properties \ident{Record} and \ident{Array}.

@@

\section{Formalizing type providers}
\label{sec:formal}

@@

\subsection{Capitalization} 
\label{sec:impl-naming}

\subsection{Predictability and stability} 
\label{sec:impl-stable}

In particular, we do 
not use probabilistic methods where adding an additional sample could completely change the 
shape of the type.
